---
title: "Practical Machine Learning Project"
author: "Achala Acharya"
date: "02/22/2015"
output: html_document
---

## Introduction

In this document I will describe, how I built a model to predict the activity class for dataset provided for Practical Machine Learning Course from the Data SCience Specialization.

The first step in this process will be downloading the data sets for use in R.
```{r}
train <- read.csv('pml-training.csv')
test <- read.csv('pml-testing.csv')
head(train)
```

Then using the createpartition function in the caret package, the train set is split into test and train datasets.
```{r}
library(caret)
inTrain <- createDataPartition(y=train$classe, p=0.6, list=FALSE)
myTraining <- train[inTrain, ]; myTesting <- train[-inTrain, ]
dim(myTraining)
dim(myTesting)
```
## Data Processing and Cleanup

The next step in building the model was to cleanup the dataset and only onclude ones that are useful for prediction purpose. The following lines of code achieve this
```{r}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
head(myDataNZV)
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt","kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt","max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm","var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm","stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm","kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm","max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm","kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell","skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell","amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm","skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm","max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm","amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm","avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm","stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]
dim(myTraining)
```

Then I removed the id variables and also removed variables that had more than 60% missing values.

```{r}
myTraining <- myTraining[,-c(1)]
trainingV3 <- myTraining 
for(i in 1:length(myTraining)) { #for every column in the training dataset
  if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { 
    for(j in 1:length(trainingV3)) {
      if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1)  { 
        trainingV3 <- trainingV3[ , -j] 
      }
    }
  }
}

dim(trainingV3)
myTraining <- trainingV3
rm(trainingV3)
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) 
myTesting <- myTesting[clean1]
testing <- test[clean2]
dim(myTesting)

for (i in 1:length(testing) ) {
  for(j in 1:length(myTraining)) {
    if( length( grep(names(myTraining[i]), names(testing)[j]) ) ==1)  {
      class(testing[j]) <- class(myTraining[i])
    }
  }
}

testing <- rbind(myTraining[2, -58] , testing) 
testing <- testing[-1,]
```

## Model Building: Random Forest Model

After a series of cleanup process the data is now ready to be trained using a machine learnig techniue. For this dataset I will fit a random forest model to predict the activity classe variable.

```{r}
library(randomForest)
fit2 <- randomForest(classe ~. , data=myTraining)
prediction2 <- predict(fit2, myTesting, type = "class")
```

After building the model I tested the performance with a confusion matrix to see how accurately the model classifies activity type.

```{r}
confusionMatrix(prediction2, myTesting$classe)
```

The confusion matrix shows that the model does effectively classifies activity type with a small error rate. 

## Generating files for submission

This model was then be used to predict the test set for the submission part of the assignment
```{r}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}
pml_write_files(prediction2)
```

